Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-30 main.tex
Adjust the line arrangement of the subcaption figures so that pandoc parses them in the html conversion.

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were the
two actions which necessitated stopping the treadmill belt and they are proxies
for real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Turning the balance assist system on significantly \((p=0.02<0.05)\) reduces
the odds that a perturbation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off. This figure is created by setting all
explanatory variables to their centered mean values and calculating the
probability from Equation~\ref{eq:log-regress} for only change in angular
impulse given the estimates in Table~\ref{tab:freq-coefs-6}. The table
indicates that the balance assist system halves (0.53) the odds that a
perturbation results in a fall.  This figure shows that for relatively large
impulses the probability to fall is unity for both states of balance assist on
and off. And for relatively small impulses the probability to fall is null for
both states. But for impulses in the magnitude region (about -1 to 0.5 STD),
i.e. around the mean-centered perturbation resistance threshold, the
probability of falling is significantly lowered with the balance assist system
on. The skewness of the probability curves arrives from the interaction
effects. Figure~\ref{fig:probability-10kph} shows the same result for the
10~\si{\kph} trials which has a similar trend of reducing the probability to
fall with the balance assist system turned on, but the effect is not
significant \((p=0.07 > 0.05)\).
%
\begin{figure}
  \centering
  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance assist on
    (solid) or off (dashed) when all other predictor variables are set to zero,
    except for the interaction effect between the balance assist and the
    angular impulse. The abscissa is the standard deviation around the mean
    of all perturbations, centered per participant.}
  \label{fig:probability}
\end{figure}

>>>>

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were the
two actions which necessitated stopping the treadmill belt and they are proxies
for real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
% NOTE : The html conversion with pandoc fails to include these subcaption
% figures if the includegraphics line is on the same line as the subcaptionbox
% line (or maybe simply if the curly braces containing the includegraphics call
% is on the same line as the subcaptionbox.
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}
    {\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}
    {\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}
    {\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}
    {\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Turning the balance assist system on significantly \((p=0.02<0.05)\) reduces
the odds that a perturbation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off. This figure is created by setting all
explanatory variables to their centered mean values and calculating the
probability from Equation~\ref{eq:log-regress} for only change in angular
impulse given the estimates in Table~\ref{tab:freq-coefs-6}. The table
indicates that the balance assist system halves (0.53) the odds that a
perturbation results in a fall.  This figure shows that for relatively large
impulses the probability to fall is unity for both states of balance assist on
and off. And for relatively small impulses the probability to fall is null for
both states. But for impulses in the magnitude region (about -1 to 0.5 STD),
i.e. around the mean-centered perturbation resistance threshold, the
probability of falling is significantly lowered with the balance assist system
on. The skewness of the probability curves arrives from the interaction
effects. Figure~\ref{fig:probability-10kph} shows the same result for the
10~\si{\kph} trials which has a similar trend of reducing the probability to
fall with the balance assist system turned on, but the effect is not
significant \((p=0.07 > 0.05)\).
%
\begin{figure}
  \centering
  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.\label{fig:probability-6kph}}
    {\includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}}
  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.\label{fig:probability-10kph}}
    {\includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}}
  \caption{Comparison of predicted fall probability for balance assist on
    (solid) or off (dashed) when all other predictor variables are set to zero,
    except for the interaction effect between the balance assist and the
    angular impulse. The abscissa is the standard deviation around the mean
    of all perturbations, centered per participant.}
  \label{fig:probability}
\end{figure}
